How Slow do my Transient Typechecks Go?

### 2019-08-17 paper.tex
Conclusion!

Although na\"ive implementations of transient type checks have a high runtime overhead, just-in-time compilation and optimisation in virtual machines can eliminate much of this overhead.  Unfortunately
the improvement is not uniform: while most type checks can be optimised
away, some will significantly decrease a program's
performance, and some may even increase it.  In this paper we investigate techniques that interpret benchmark results to identify which type checks have the greatest effects on performance.
Using these techniques,
programmers can optimise programs by removing expensive type checks,
and VM engineers can identify new opportunities for compiler optimisation.

@@

By inspecting the scatterplots, we observe that the performance of
Moth in almost any configuration of a benchmark is bounded by
the performance in the untyped and fully-typed configuration.
That is, for these benchmarks on transient type checks,
measuring just the untyped and fully-typed configurations
would provide excellent estimates of a benchmark's
performance bounds.  This is different to the experience
of other kinds of gradual typing, where the best, and most
importantly worst, configurations are not always those
fully typed or fully untyped \cite{Greenman2019jfp}.
%
However, the Richards and Snake benchmarks \textit{do} have
sections outside these bounds, and isolated executions of
a few others (SpectralNorm, JSON, DeltaBlue) are also outliers.
%
We believe that the 0\% and 100\% bounds are nonetheless a
reasonable heuristic estimator in most cases, and will
examine further the outlying benchmarks.

Of particular note is that some of the graphs (Permute, Storage, Snake, Towers, and Richards) appear to show bimodal performance profiles, that is, two separate roughly-horizontal lines. Presumably Moth can remove all the overhead from some configurations, but in others there must be one or more type checks that cannot be optimised away. List appears to have three performance modes: most of the configurations up to about 60\% annotations perform at roughly the same speed as untyped code; most configurations above 70\% perform 1.8 times more slowly than untyped code; and then across all annotation percentages there are configurations that perform roughly 1.5 times slower than the untyped baseline.

@@

In this paper we have investigated how benchmark results can be
repurposed to determine precisely which transient typechecks are
resistant to the optimisations provided by a just-in-time compiling
virtual machine.  We observed that many of our benchmark results
conducted under the Takikawa protocol 
seemed to have a characteristic bimodal performance profile: some
of the benchmark configurations ran significantly slower than
the remaining configurations. By inspecting the benchmark graphs coded
for each typecheck in a program's source code, we were often able to
identify that just one or two of the typechecks were responsible for
the bimodal performance.  Based on that result, we also investigated
individual benchmark configurations for each individual typecheck:
these results can identify problematic typechecks, in some but not all cases.

This is preliminary work: there are a number of threats to validity. 
Regarding construct validity, 
our underlying
implementation may contain undetected bugs that affect the semantics
or performance of the gradual typing checks. Regarding internal
validity,
our benchmarking harness runs on the same implementation
and therefore is subject to the same issues.
Regarding external validity, 
\RR{The next two sentences seem to be in the wrong place (they state an advantage rather than a threat); maybe move them into the previous paragraph?}
\kjx{I'm happy with them here about external validity}
Moth is built
on the Truffle and Graal toolchain, so we expect
to resemble other Graal
VMs doing similar AST-based optimizations of transient typechecks.
Because we rely on common techniques, 
we expect our results should be transferable to JIT implementations
more widely.

Finally, it is not clear how our results would transfer
to other gradually typed-languages or other semantics for gradual typing.
Our benchmarks do not depend on any features of Grace
that are not common in other gradually-typed object-oriented
languages, but as Grace lacks a large corpus of programs the
benchmarks are necessarily small and artificial.
The advantage of Grace and Moth for this research is
that their relative simplicity means we have been able to build an
implementation that features competitive performance with significantly less
effort than would be required for larger and more complex languages.

In the future, we hope to investigate statistical techniques to
determine the significance of each type annotation's contribution to a
programs overall performance. We would also like to investigate 
whether this approach can assist with optimsations for programmers' day-to-day  development, or help VM engineers identifying performance bugs in the underlying virtual machines.

\sm{is our approach practical. Is it useful for day-to-day development?
What's its context, use, general application?}
\kjx{filled out the last sentence}

>>>>

Although na\"ive implementations of transient type checks have a high runtime overhead, just-in-time compilation and optimisation in virtual machines can eliminate much of this overhead.  Unfortunately
the improvement is not uniform: while most type checks can be optimised
away, some will significantly decrease a program's
performance, and some may even increase it.  In this paper we show techniques where benchmark results can be used to help identify which type annotations have the greatest effects on performance.
Using these techniques,
programmers can optimise programs by removing expensive type checks,
and VM engineers can identify new opportunities for compiler optimisation.

@@

Of particular note is that some of the graphs (Permute, Storage, Snake, Towers, and Richards) appear to show bimodal performance profiles, that is, two separate roughly horizontal lines. Presumably Moth can remove all the overhead from some configurations, but in others there must be one or more type checks that cannot be optimised away. List appears to have three performance modes: most of the configurations up to about 60\% annotations perform at roughly the same speed as untyped code; most configurations above 70\% perform 1.8 times more slowly than untyped code; and then across all annotation percentages there are configurations that perform roughly 1.5 times slower than the untyped baseline.

@@

In this paper we have investigated how benchmarks can be
repurposed to determine precisely which transient type checks are
resistant to the optimisations provided by a just-in-time compiling
virtual machine.  We observed that many of our benchmark results
conducted under the Takikawa protocol 
seemed to have a characteristic bimodal performance profile: some
of the benchmark configurations ran significantly slower than
the remaining configurations. We also observed trimodal profiles, as well as performance increases when type annotations where added.
By inspecting benchmark graphs colour coded for each individual type annotation in a program's source code, we were often able to
identify that just one or two of the type checks were likely responsible for
the bimodal performance, by observing patterns in the associated graph.
However, this approach requires inspecting one graph per type annotation.

We also investigated another approach to help identify such annotation, for an individual benchmark, we can produce a single graph of the performances of individual configurations for each type annotation. 
In most cases both approaches identified the same type annotations as likely causing a significant effect on performance.

This is preliminary work, in particular we have not yet identified exactly why these type annotations appear to have an effect on performance.

There are also a number of threats to validity. 
Regarding construct validity, 
our underlying
implementation may contain undetected bugs that affect the semantics
or performance of the gradual typing checks. Regarding internal
validity,
our benchmarking harness runs on the same implementation
and therefore is subject to the same issues.
Regarding external validity, 
%\RR{The next two sentences seem to be in the wrong place (they state an advantage rather than a threat); maybe move them into the previous paragraph?}
%\kjx{I'm happy with them here about external validity}
Moth is built
on the Truffle and Graal toolchain, so we expect
to resemble other Graal
VMs doing similar AST-based optimizations of transient type checks.
Because we rely on common techniques, 
we expect our results to be transferable to other JIT implementations as well.

Finally, it is not clear how our results would transfer
to other gradually typed-languages or other semantics for gradual typing.
Our benchmarks do not depend on any features of Grace
that are not common in other object-oriented
languages, but as Grace lacks a large corpus of programs the
benchmarks are necessarily small and artificial.
The advantage of Grace for this research is
that their relative simplicity means we have been able to build an
implementation that features competitive performance with significantly less
effort than would be required for larger and more complex languages.

In the future, we hope to investigate statistical techniques to
determine the significance of each type annotation's contribution to a
programs overall performance. We would also like to investigate 
whether this approach can assist with optimsations for programmers' day-to-day  development, or help VM engineers identifying performance bugs in the underlying virtual machines.
\sm{is our approach practical. Is it useful for day-to-day development?
What's its context, use, general application?}
\kjx{filled out the last sentence}
